Modelling Prehistoric Iconographic Compositions. The \proglang{R} package \pkg{decorr}

### 2020-12-13 article_rvTH17.Rnw
version17

%% - \author{} with primary affiliation
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{Thomas Huet\\UMR 5140 \And Craig Alexander \And Jose Pozo}
%   \And Second Author\\Plus Affiliation}
\Plainauthor{Thomas Huet, Craig Alexander, Jose Pozo}

@@

Furthermore, at the graphical units level, top-down hierachical typologies (e.g., a category of graphical units with \emph{n} types, a type with \emph{n} subtypes, a subtype with \emph{n} varieties, etc.) lead also to tedious inventories and difficult the comparisons \emph{within} and \emph{between} the categories.

@@

According to us, prehistorical composition -- like any formal system -- are spatial features (mostly shapes) related one with the other depending on rules of spatial proximities. Then, the First Law of Geography: "\emph{everything is related to everything else, but near things are more related than distant things}" \citep{Tobler70} should be more operative than the Saussurian '\emph{linearity of the signifier}' principle \citep{Saussure89}. To get these '\emph{near things}', or features proximities, we use concepts coming from the Graph Theory, which offers a general framework to manage neighbourhoods, and particularly relative neighborhood graphs (RNG) which is '\emph{a powerful model of low-level visual process}' \citep{Toussaint80}. Furthermore, we address the questions of managing graphical units attributes (for example an anthropomorph with a sword \emph{vs} an anthropomorph with a spear) and superimpositions of graphical units.

@@

Graph theory offers a conceptual framework and indices (global at the entire graph scale, local at the vertex scale) to deal with notions of networks, relationships and neighbourhoods. Graphical units (GUs) can be modelled as vertices (nodes) separated one with another by undecorated areas. Their spatial relations can be modelled as edges. The different levels of GUs spatial organisation can be retrieve by a relative neighborhood graphs analysis (Graph Theory) and a spatial (GIS) analysis.

Graph elements, nodes and edges -- repectively GUs and connexions between GUs -- are created on a GIS interface. Indeed, for large series of decorated supports, GIS offers the most suitable and flexible interface to register all GUs and to get their coordinates. The decoration image is considerated as the basemap of the project and will cover the region of interest of the analysis (ie, all the graphical content). The decoration image can be binarized: GUs are considerated active, the undecorated parts of the support -- the background -- are considerated inactive. The polygonisation of the GUs (\code{POLYGONS}), including the border of the stelae, allows to calculated their Voronoi seeds (ie, their centroid) and Voronoi cells (ie, their area of influence). So, the entire decoration image is tiled and condidered as a Voronoi diagram. Centroids of the GUs (\code{POINTS}) are calculated and considered as the graph nodes, their locations (\emph{x} and \emph{y} coordinates, measured in pixels) are relative to each decoration. Exist edges (\code{LINES}) between pairs of GUs when their Voronoi cells share a common border (\emph{birel}: 'touches').

The entire graphical content of the decoration is mapped with a neighborhood graphs (RNG) -- ie geometric graph, or planar graph -- of GUs centroids (nodes) and GUs proximity links (edges). So, a decoration graph is a one component graph. This graph is the dual graph of the decoration Voronoi diagram (Fig. \ref{fig:gis}).

@@

This model has a minimal of \emph{a priori} definitions. Those definitions only concern the graph elements: nodes variables (ie, GUs variables) and types of relations they share (edges).

@@

\begin{figure}[H]
% \begin{figure}[!htbp]
\centering\emph{}
\includegraphics[height=10cm, trim=15mm 20mm 15mm 20mm, clip, keepaspectratio]{gis_1}
\caption{\label{fig:gis_1} GIS interface, Cerro Muriano 1 decoration. The nodes \code{casque} (helmet) and \code{sexe\_masculin} (male sex) are two nodes attributes of the node \code{personnage} (anthropomorph).}
\end{figure}

Broadly speaking, for further statistical analysis, it is better to use this \emph{attribute} method than to multiply the categories of nodes.

\item[\emph{overlapping edges}]
Finally, it occurs that a graphical composition shows decoration phases. These decoration phases can be identified by observing the superimpositions between different nodes or by other means (the technique or the patina for example). In the first case, two nodes belonging to two different phase will have a common part (the overlapped intersection) while in the second case the two nodes will not necessarly share a part of their surface. To record this relative chronology, an edge between two non-contemporaneous nodes is tagged \code{'>'} (\code{type = '>'}), represented with a blue plain line and its textual notation is \code{'->-'}. For example \code{ecriture ->- bouclier} means that \code{ecriture} is more recent than the \code{bouclier}, or \emph{overlaps} the\code{bouclier} in the stratigraphical sense. The edge is directed since \code{ecriture ->- bouclier} (ie, \code{ecriture} \emph{overlaps} \code{bouclier} ) is not equal to \code{bouclier ->- ecriture} (ie, \code{bouclier} \emph{overlaps} \code{ecriture}) (Fig. \ref{fig:gis_2}).

@@

\begin{figure}[H]
% \begin{figure}[!htbp]
\centering
\includegraphics[height=10cm, trim=30mm 16mm 30mm 13mm, clip, keepaspectratio]{gis_2}
\caption{\label{fig:gis_2} GIS interface, Ibahernando decoration. The node \code{ecriture} (writing) overlaps \code{lance} (spear) and the upper part of the \code{bouclier} (shield).}
\end{figure}

@@

These different relationship types are resumed with their birel equivalencies in Table \ref{table:tedges}.

\begin{table}[!htbp]
\centering
\begin{tabular}{ccccp{2cm}p{4.9cm}}
\hline
% node A & type of edge & node B & undirected & birel & stratigraphical meaning \\
Node 1 & Edge type & Node 2 & Edge directness & \emph{Birel} & Description \rule{0pt}{1em} \\[1pt]
\hline \rule{0pt}{1em}%
\code{A} & \code{=} & \code{B} & Undirected & $ A \cap B = \emptyset $ & 
\code{A} and \code{B} are contemporaneous and disjoint. \\
\code{A} & \code{+} & \code{B} & Directed & $ A \cap B = A $ & 
\code{A} and \code{B} are contemporaneous. \code{B} is an attribute of \code{A}. \\
\code{A} & \code{>} & \code{B} & Directed & $ A \cap B = \exists $ \newline or \newline $ A \cap B = \emptyset $ & 
\code{A} is more recent than \code{B}. \\
% \code{1} & = & 2 & undirected & A \cap B = \varnothing & 1 and 2 can be contemporaneous \\
\hline
\end{tabular}
\caption{Synthesis for the different types of relations between GUs.}\label{tab1}
\label{table:tedges}
\end{table}

@@

A training dataset (nodes and edges coordinates, decoration images) is stored in the \code{extdata} folder of the \pkg{decorr} package
\smallbreak
% \begin{itemize}
% \item The \code{imgs} dataframe
% \end{itemize}
\textbf{\code{imgs} dataframe}
\\
The inventory of decorations is stored in the \code{imgs} dataframe (Tab. \ref{tab:timg}). The field \code{imgs$idf} is the short name of the decoration, useful during statistical analysis. The primary key of each decoration is the concatenate of \code{imgs$site} and \code{imgs$decor}. These keys will allow joints with the other dataframes (\code{nodes} and \code{edges})
%
<<results='asis', echo=FALSE, warning=FALSE>>=
imgs <- read.table(system.file("extdata", "imgs.tsv", package = "decorr"),
                   sep="\t", stringsAsFactors = FALSE)
print(xtable::xtable(imgs,
                     caption="The studied corpus, the \\code{imgs.tsv} dataframe",
                     label="tab:timg",
                     size=6,
                     align=c(
                       "p{0.01cm}",
                       "c",
                       ">{\\raggedright\\arraybackslash\\rule{0pt}{1em}\\addtolength{\\baselineskip}{-1.5pt}}p{2.1cm}",
                       ">{\\raggedright\\arraybackslash\\addtolength{\\baselineskip}{-1.5pt}}p{2.1cm}",
                       "p{8.5cm}")),
      table.placement="!htbp",
      include.rownames=FALSE)
@
%
The \pkg{decorr} package training dataset is composed by five stelaes decorations drawings \citep{DiazGuardamino10} belonging to the so-called 'Warrior stelae' family -- with about 140 stelae -- dated to the Late Bronze Age of SW Iberian peninsula \citep{Celestino01}. \\
At first the drawings dataset can be checked by using the \code{imgs} dataframe and the \pkg{magick} package (Fig. \ref{fig:dataset})
%
<<timg, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=15, fig.height=11.2, fig.cap="\\label{fig:dataset}Decoration images of the training dataset">>=
pth <- system.file("extdata", package = "decorr")
imgs <- read.table(system.file("extdata", "imgs.tsv", package = "decorr"),
                   sep="\t", stringsAsFactors = FALSE)
lims <- list()
for(i in 1:nrow(imgs)){
  i1 <- image_read(paste0(pth,"\\",imgs[i,"img"]))
  lbl.txt <- paste0(imgs[i,"idf"],"\n",
                    imgs[i,"site"],"\n",
                    imgs[i,"decor"],"\n",
                    imgs[i,"img"],"\n",
                    image_info(i1)$width,"*",image_info(i1)$height," px")
  i1 <- image_annotate(i1,lbl.txt,location = "northwest",
                       size = 28, color = "red")
  lims[[length(lims)+1]]<- i1
}
out.img <- image_append(c(image_append(c(lims[[1]],lims[[2]],lims[[3]])),
                          image_append(c(lims[[4]],lims[[5]]))),
                        stack = TRUE)
par(mar=c(0,0,0,0))
plot(out.img)
@
%
To construct a graph overlapping the decoration images listed in the \code{images} dataframe, the first step is to load \code{nodes} and \code{edges} dataframes.
% \begin{itemize}
% \item The \code{nodes} dataframe
% \end{itemize}
\smallbreak
\textbf{Nodes dataframe}
\\
It contains the required minimum variables for the analysis (Tab. \ref{tab:tnds}).

<<tndss, results='asis', echo=TRUE>>=
caption <- "Nodes (from \\code{nodes.csv} dataframe)"
nodes <- read.table(system.file("extdata", "nodes.csv", package = "decorr"),
                    sep=";",stringsAsFactors = FALSE)
print(xtable::xtable(head(nodes),
                     caption = caption,
                     label="tab:tnds"),
      table.placement="!htbp")
@

Since a prehistorical site can have various decorated objects, the primary key of the decoration is based on two fields: \code{nodes$site} and \code{nodes$decor}. The \code{nodes$id} is the identifier of the node. The \code{nodes$type} field is the default variable for further statistical analysis. Here, \code{nodes$type} refers to the typology of the nodes (anthropomorph, weapons, etc.). The \code{nodes$x} and \code{nodes$y} columns refer to the \emph{x} and \emph{}y coordinates of the nodes. As said, in the first place theses coordinates come from the GIS. But, in a GIS, the coordinates origin (0, 0) is the bottom-left corner and exist negative values, while this origin is top-left for any \proglang{R} matrices (rasters, grids, dataframes, etc.) with only positive values. To recover the local \emph{y} value of nodes and edges on the decoration image, the absolute value \emph{y} value and the image height, as a constant offset, are computed.
\smallbreak
% \begin{itemize}
% \item The \code{edges} dataframe
% \end{itemize}
\textbf{\code{edges} dataframe}
\\
The \code{edges} dataframe is quite similar to the \code{nodes} dataframe (Table \ref{tab:teds}).

<<tedss, results='asis', echo=TRUE>>=
caption <- "Edges of Cerro Muriano 1 (from \\code{edges.csv} dataframe)"
edges <- read.table(system.file("extdata", "edges.csv", package = "decorr"),
                    sep=";",stringsAsFactors = FALSE)
edges.cm1 <- subset(edges, decor == "Cerro Muriano 1")
print(xtable::xtable(head(edges.cm1),
                     caption=caption,
                     label="tab:teds"),
      table.placement="!htbp")
@

Fields \code{edges$site} and \code{edges$decor} are the primary key of decoration. The fields \code{edges$a} and \code{edges$b} are the equivalent to columns \emph{from} and \emph{to} in Graph theory. Even if undirected graphs will the most common in further studies, this direction helps to distinguish between nodes. The column \code{edges$a} is the identifier of \emph{starting node} or \emph{main node} or \emph{overlapping node}. The column \code{edges$b} is the identifier of the \emph{ending node} or \emph{attribue node} or \emph{overlapped node}. The column \code{edges$type} is the type of relation (normal, attribute, overlapping, etc.) between the node a and the node b. There is no need to record the coordinates of the edges, these coordinates are calculated from the \code{nodes} dataframe.

The \pkg{decorr} package functions can be divided into:
\begin{enumerate}
\item management functions
\item graphical functions
\item single decoration functions
\item comparisons between different decorations functions
\end{enumerate}

@@

\end{Begin}
list_dec(imgs,
         nodes,
         edges)
\end{Code}

Once \code{imgs}, \code{nodes} and \code{edges} dataframes loaded, the \code{list\_dec()} function allows to create graphs for all decorations within these dataframes. The result is a list of \code{igraph} objects. As an example, the first graph of the list can be plotted with the \code{igraph::plot()} function (Fig. \ref{fig:igraph})

<<igraphs, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=2.5, fig.height=2.5, fig.cap="\\label{fig:igraph}R interface. Plot of the first graph of the list (Cerro Muriano 1 decoration)">>=
lgrph <- list_dec(imgs, nodes, edges)
par(mar=c(0, 0, 0, 0))
plot(lgrph[[1]],
     vertex.color = "orange",
     vertex.frame.color = "orange",
     vertex.label.color = "black",
     vertex.size = 16,
     vertex.label.cex = 1,
     edge.color = "orange",
     vertex.label.family = "Helvetica"
)
@

@@

<<named, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=2.5, fig.height=2.5, fig.cap="\\label{fig:igraph}R interface. Plot of the first graph of the list (Cerro Muriano 1 decoration)">>=
lgrph <- list_dec(imgs, nodes, edges)
as.character(named_elements(lgrph[[2]], focus = "edges", nd.var="type"))
as.character(named_elements(lgrph[[2]], focus = "nodes", nd.var="type"))
@

@@

The \pkg{decorr} package has two graphical functions.

@@

Functions allowing to create a RNG for a single decoration are \code{read\_nds()} and \code{read\_eds()}. These functions allow to read respectively a file of nodes and a file of edges (\code{.csv}, \code{.tsv} or \code{.shp} files).

@@

\code{read_nds()} is close to the \proglang{R} native \code{read.table()} function, but allows to read \emph{shapefiles} of nodes. For example, for the Torrejon Rubio 1 decoration, the node 6 is a sword (\code{epee}) (Table \ref{tab:nd}).

<<nd, results='asis', echo=TRUE>>=
caption <- "Torrejon Rubio 1 decoration node 6: the sword"
sit <- "Torrejon Rubio" ; dec <- "Torrejon Rubio 1"
nds.df <- read_nds(site = sit, decor = dec, dev = ".shp",
                   doss = system.file("extdata", package = "decorr"))
print(xtable::xtable(nds.df[nds.df$id == 6, ],
                     caption = caption,
                     label = "tab:nd",
                     size = 7),
      table.placement = "!htbp")
@

@@

\code{read_eds()} permits to read a \emph{shapefiles} of edges (\code{dev = ".shp"}) or, when the data source is a \code{.csv} or \code{.tsv} dataframe, to retrieve the coordinates of the edges from the \code{nodes} dataframe (Table \ref{tab:ed}).

>>>>

%% - \author{} with primary affiliation
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{Thomas Huet\\UMR 5140 \And Jose Pozo \And Craig Alexander}
%   \And Second Author\\Plus Affiliation}
\Plainauthor{Thomas Huet, Jose Pozo, Craig Alexander}

@@

@@

According to us, prehistorical composition -- like any formal system -- are spatial features (mostly shapes) related one with the other depending on rules of spatial proximities. Then, the First Law of Geography: "\emph{everything is related to everything else, but near things are more related than distant things}" \citep{Tobler70} should be more operative than the Saussurian '\emph{linearity of the signifier}' principle \citep{Saussure89}. To get these '\emph{near things}', or features proximities, we use concepts coming from the Graph Theory, which offers a general framework to manage neighbourhoods. Furthermore, we address the questions of managing graphical units attributes (for example an anthropomorph with a sword \emph{vs} an anthropomorph with a spear) and superimpositions of graphical units.  
% and particularly relative neighborhood graphs (RNG) which is '\emph{a powerful model of low-level visual process}' \citep{Toussaint80}

@@

Graph theory offers a conceptual framework and indices (global at the entire graph scale, local at the vertex scale) to deal with notions of networks, relationships and neighbourhoods. Graphical units (GUs) can be modelled as vertices (nodes) separated one with another by undecorated areas. Their spatial relations can be modelled as edges. The different levels of GUs spatial organisation can be retrieve by graphs analysis (Graph Theory) and a spatial (GIS) analysis.

Graph elements, nodes and edges -- repectively GUs and connexions between GUs -- are created on a GIS interface.  
Indeed, for large series of decorated supports, GIS offers the most suitable and flexible interface to register all GUs and to get their coordinates, makes easy the updating of GUs and GUs relation and opens to develop further spatial analysis.  The decoration image is considerated as the basemap of the project and will cover the region of interest of the analysis (ie, all the graphical content). The decoration image can be binarized: GUs are considerated active, the undecorated parts of the support -- the background -- are considerated inactive. The polygonisation of the GUs (\code{POLYGONS}), including the border of the stelae, allows to calculated their Voronoi seeds (ie, their centroid) and Voronoi cells (ie, Thiessen polygons, their area of influence). So, the entire decoration image is tiled and condidered as a Voronoi diagram. Centroids of the GUs (\code{POINTS}) are calculated and considered as the graph nodes, their locations (\emph{x} and \emph{y} coordinates, measured in pixels) are relative to each decoration. Exist edges (\code{LINES}) between pairs of GUs when their Voronoi cells share a common border (\emph{birel}: 'touches').

The entire graphical content of the decoration is mapped with a geometric graph -- or planar graph -- of GUs centroids (nodes) and GUs proximity links (edges). So, a decoration graph is a one component graph. This graph is the dual graph of the decoration Voronoi diagram (Fig. \ref{fig:gis}).

@@

@@

% \begin{figure}[H]
\begin{figure}[!htbp]
\centering\emph{}
\includegraphics[height=7cm, trim=15mm 20mm 15mm 20mm, clip, keepaspectratio]{gis_1}
% \includegraphics[height=10cm, trim=15mm 20mm 15mm 20mm, clip, keepaspectratio]{gis_1}
\caption{\label{fig:gis_1} GIS interface, Cerro Muriano 1 decoration. The nodes \code{casque} (helmet) and \code{sexe\_masculin} (male sex) are two nodes attributes of the node \code{personnage} (anthropomorph).}
\end{figure}

\item[\emph{overlapping edges}]
Finally, it occurs that a graphical composition shows decoration phases. These decoration phases can be identified by observing the superimpositions between different nodes or by other means (differences in the techniques or in the patina for example). In the first case, two nodes belonging to two different phase will have a common part (the overlapped intersection) while in the second case the two nodes will not necessarly share a part of their surface. To record this relative chronology, an edge between two non-contemporaneous nodes is tagged \code{'>'} (\code{type = '>'}), represented with a blue plain line and its textual notation is \code{'->-'}. For example \code{ecriture ->- bouclier} means that \code{ecriture} is more recent than the \code{bouclier}, or \emph{overlaps} the\code{bouclier} in the stratigraphical sense. The edge is directed since \code{ecriture ->- bouclier} (ie, \code{ecriture} \emph{overlaps} \code{bouclier}) is not equal to \code{bouclier ->- ecriture} (ie, \code{bouclier} \emph{overlaps} \code{ecriture}) (Fig. \ref{fig:gis_2}).

@@

\begin{figure}[H]
% \begin{figure}[!htbp]
\centering
\includegraphics[height=8cm, trim=30mm 16mm 30mm 13mm, clip, keepaspectratio]{gis_2}
\caption{\label{fig:gis_2} GIS interface, Ibahernando decoration. The node \code{ecriture} (writing) overlaps \code{lance} (spear) and the upper part of the \code{bouclier} (shield).}
\end{figure}

@@

The nodes and edges equalities are resumed in the Table \ref{tab:ex.comm}.

<<equalities, results='asis', echo=FALSE, fig.pos = '!htbp'>>=
caption <- paste0("Example of equalities and differences for nodes and edges ")
df.ex <- data.frame(type = c("node", "node", "edge", 
                             "edge", "edge", "edge", "edge"),
                    part.1 = c("epee", "epee", "epee --=-- lance",
                               "epee --=-- lance", "personnage --=-- bouclier", 
                               "personnage --+-- bouclier", "ecriture -->-- bouclier"),
                    part.2 = c("epee", "lance", "epee --=-- lance",
                               "lance --=-- epee", "personnage --+-- bouclier",
                               "bouclier --+-- personnage", "bouclier -->-- ecriture"),
                    is.equal = c("True", "False", "True", 
                                 "False", "False", "False", "False"),
                    stringsAsFactors = F)
print(xtable::xtable(df.ex,
                     caption = caption,
                     label = "tab:ex.comm",
                     size = 7),
      include.rownames = FALSE,
      table.placement = "!htbp")
@

These different relationship types are resumed with their \emph{birel} equivalencies and chronological assumptions in Table \ref{table:tedges}.

% \begin{table}[!htbp]
\begin{table}[H]
\centering
\begin{tabular}{ccccp{2cm}p{4.9cm}}
\hline
% node A & type of edge & node B & undirected & birel & stratigraphical meaning \\
Node 1 & Edge type & Node 2 & Edge directness & \emph{Birel} & Description \rule{0pt}{1em} \\[1pt]
\hline \rule{0pt}{1em}%
\code{A} & \code{=} & \code{B} & Undirected & $ A \cap B = \emptyset $ & 
\code{A} and \code{B} are contemporaneous and disjoint. \\
\code{A} & \code{+} & \code{B} & Directed & $ A \cap B = A $ & 
\code{A} and \code{B} are contemporaneous. \code{B} is an attribute of \code{A}. \\
\code{A} & \code{>} & \code{B} & Directed & $ A \cap B = \exists $ \newline or \newline $ A \cap B = \emptyset $ & 
\code{A} is more recent than \code{B}. \\
% \code{1} & = & 2 & undirected & A \cap B = \varnothing & 1 and 2 can be contemporaneous \\
\hline
\end{tabular}
\caption{Synthesis for the different types of relations between GUs.}\label{tab1}
\label{table:tedges}
\end{table}

@@

A training dataset is stored in the \code{extdata} folder of the \pkg{decorr} package. It contains dataframes of nodes' coordinates, edges' coordinates, decorations' listing in different formats (\code{.tsv}, \code{.csv} or \emph{shapefiles}) and the decoration drawings (\code{.jpg}) of five Late Bronze Age stelae belonging to the so-called 'Warrior stelae' family -- with about 140 stelae -- mostly concentrated in the Southwest Iberian peninsula \citep{Celestino01}. \\

\smallbreak
\textbf{Imgs dataframe}
\\
The inventory of decorations is stored in the \code{imgs} dataframe (Tab. \ref{tab:timg}). The field \code{imgs$idf} is the short name of the decoration, useful during statistical analysis. The primary key of each decoration is the concatenate of \code{imgs$site} and \code{imgs$decor}. These keys will allow joints with the other dataframes (nodes and edges)
%
<<results='asis', echo=FALSE, warning=FALSE>>=
imgs <- read.table(system.file("extdata", "imgs.tsv", package = "decorr"),
                   sep="\t", stringsAsFactors = FALSE)
print(xtable::xtable(imgs,
                     caption = "The studied corpus, the \\code{imgs.tsv} dataframe",
                     label = "tab:timg",
                     size=6,
                     align=c(
                       "p{0.01cm}",
                       "c",
                       ">{\\raggedright\\arraybackslash\\rule{0pt}{1em}\\addtolength{\\baselineskip}{-1.5pt}}p{2.1cm}",
                       ">{\\raggedright\\arraybackslash\\addtolength{\\baselineskip}{-1.5pt}}p{2.1cm}",
                       "p{8.5cm}")),
      table.placement="!htbp",
      include.rownames=FALSE)
@
%

At first the drawings dataset \citep{DiazGuardamino10} can be checked by using the imgs dataframe and the \pkg{magick} package (Fig. \ref{fig:dataset})
%
<<timg, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=15, fig.height=11.2, fig.cap="\\label{fig:dataset}Decoration images of the training dataset">>=
pth <- system.file("extdata", package = "decorr")
imgs <- read.table(system.file("extdata", "imgs.tsv", package = "decorr"),
                   sep="\t", stringsAsFactors = FALSE)
lims <- list()
for(i in 1:nrow(imgs)){
  i1 <- image_read(paste0(pth,"\\",imgs[i,"img"]))
  lbl.txt <- paste0(imgs[i,"idf"], "\n",
                    imgs[i,"site"], "\n",
                    imgs[i,"decor"], "\n",
                    imgs[i,"img"], "\n",
                    image_info(i1)$width, "*", image_info(i1)$height, " px")
  i1 <- image_annotate(i1, lbl.txt, location = "northwest",
                       size = 28, color = "red")
  lims[[length(lims)+1]]<- i1
}
out.img <- image_append(c(image_append(c(lims[[1]], lims[[2]], lims[[3]])),
                          image_append(c(lims[[4]], lims[[5]]))),
                        stack = TRUE)
par(mar=c(0, 0, 0, 0))
plot(out.img)
@
%
To construct a graph overlapping the decoration images listed in the \code{images} dataframe, the first step is to load nodes and edges dataframes.
% \begin{itemize}
% \item The \code{nodes} dataframe
% \end{itemize}
\smallbreak
\textbf{Nodes dataframe}
\\
It contains the required minimum variables for the analysis (Tab. \ref{tab:tnds}).

<<tndss, results='asis', echo=TRUE>>=
caption <- "Nodes (from \\code{nodes.csv} dataframe)"
nodes <- read.table(system.file("extdata", "nodes.csv", package = "decorr"),
                    sep=";", stringsAsFactors = FALSE)
print(xtable::xtable(head(nodes),
                     caption = caption,
                     label = "tab:tnds"),
      include.rownames = FALSE,
      table.placement = "!htbp")
@

Since a prehistorical site can have various decorated objects, the primary key of the decoration is based on two fields: \code{nodes$site} and \code{nodes$decor}. The \code{nodes$id} is the identifier of the node. The \code{nodes$type} field is the default variable for further statistical analysis. Here, \code{nodes$type} refers to the typology of the nodes (anthropomorph, weapons, etc.). The \code{nodes$x} and \code{nodes$y} columns refer to the \emph{x} and \emph{}y coordinates of the nodes. As said, in the first place theses coordinates come from the GIS. But, in a GIS, the coordinates origin (0, 0) is the bottom-left corner and exist negative values, while this origin is top-left for any \proglang{R} matrices (rasters, grids, dataframes, etc.) with only positive values. To recover the local \emph{y} value of nodes and edges on the decoration image, the absolute value \emph{y} value and the image height, as a constant offset, are computed.
\smallbreak
% \begin{itemize}
% \item The \code{edges} dataframe
% \end{itemize}
\textbf{Edges dataframe}
\\
The edges dataframe is quite similar to the nodes dataframe (Table \ref{tab:teds}).

<<tedss, results='asis', echo=TRUE>>=
caption <- "Edges of Cerro Muriano 1 (from \\code{edges.csv} dataframe)"
edges <- read.table(system.file("extdata", "edges.csv", package = "decorr"),
                    sep=";",stringsAsFactors = FALSE)
edges.cm1 <- subset(edges, decor == "Cerro Muriano 1")
print(xtable::xtable(head(edges.cm1),
                     caption=caption,
                     label="tab:teds"),
      include.rownames = FALSE,
      table.placement="!htbp")
@

Fields \code{edges$site} and \code{edges$decor} are the primary key of decoration. The fields \code{edges$a} and \code{edges$b} are the equivalent to columns \emph{from} and \emph{to} in Graph theory. Even if undirected graphs will the most common in further studies, this direction helps to distinguish between nodes. The column \code{edges$a} is the identifier of \emph{starting node} or \emph{main node} or \emph{overlapping node}. The column \code{edges$b} is the identifier of the \emph{ending node} or \emph{attribue node} or \emph{overlapped node}. The column \code{edges$type} is the type of relation (normal, attribute, overlapping, etc.) between the node a and the node b. There is no need to record the coordinates of the edges, these coordinates are calculated from the nodes dataframe.

The \pkg{decorr} package functions can be divided into:
\begin{enumerate}
\item management functions
\item graphical functions
\item single decoration functions
\item comparisons functions
\end{enumerate}

@@

The management functions are \code{list\_dec()}, \code{contemp\_nds()} and \code{named\_elements()}

\begin{Code}
list_dec(imgs,
         nodes,
         edges)
\end{Code}

Once imgs, nodes and edges dataframes loaded, the \code{list\_dec()} function allows to create graphs for all decorations within these dataframes. The result is a list of \code{igraph} objects. As an example, the first graph of the list can be plotted with the \code{igraph::plot()} function (Fig. \ref{fig:igraph})

<<igraphs, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=2.5, fig.height=2.5, fig.cap="\\label{fig:igraph}R interface. Plot of the first graph of the list (Cerro Muriano 1 decoration)">>=
lgrph <- list_dec(imgs, nodes, edges)
par(mar=c(0, 0, 0, 0))
plot(lgrph[[1]],
     vertex.color = "orange",
     vertex.frame.color = "orange",
     vertex.label.color = "black",
     vertex.size = 14,
     vertex.label.cex = .8,
     edge.color = "orange",
     vertex.label.family = "Helvetica"
)
@

It occurs that some nodes are non-contemporaneous one with the others, like for the Ibahernando stelae. This stelae have been found reused as a funerary steale during Roman times \citep{Almagro66b}. A latin inscription (\code{ecriture}, node 1) have been carved over a spear (\code{lance}, node 2) and overlaps partially a V-notched shield (\code{bouclier}, node 3). The edges between node 1 and node 2, and the edge between node 1 and node 3, are \emph{overlap} edges (Figure \ref{fig:ibah}).

<<ibah, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=4, fig.asp=1, fig.cap="\\label{fig:ibah}Ibahernando decoration, node 1 overlaps node 2 and node 3">>=
par(mar=c(0,0,0,0))
nds.df <- read_nds(site = "Ibahernando", 
                   decor = "Ibahernando",
                   doss = system.file("extdata", 
                                      package = "decorr"))
eds.df <- read_eds(site = "Ibahernando",
                   decor = "Ibahernando",
                   doss = system.file("extdata", 
                                      package = "decorr"))
img.graph <- plot_dec_grph(nodes = nds.df,
                           edges = eds.df,
                           site = "Ibahernando",
                           decor = "Ibahernando",
                           lbl.size = 2,
                           doss = system.file("extdata", 
                                              package = "decorr"))
plot(magick::image_read(img.graph))
@

In this case, the non-contemporaneous layers of decoration should be removed before the comparison process. To do so, the original graph (1-component) will be splitted into different sub-graphs (\emph{n}-component) by selecting \code{>} edges. The studied graph component will be retrieved with the component membership of a selected node in the \code{contemp_nds()} function parameters.

To study only the Late Bronze Age iconographic layer of the Ibahernando steale, we can choose the Late Bronze Age node 4, the figure of sword (\code{epee}) dated to the middle and final stages of Late Bronze Age (ca 1250-950 BC). This node is supposed to be contemporaneous to node 2 and node 3 (Figure \ref{fig:ibah2}).

<<ibah2, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=8, fig.asp=1, fig.cap="\\label{fig:ibah2}Ibahernando stelae before and after the selection of node 4 (sword) graph component">>=
par(mar=c(0,0,0,0))
selected.nd <- 4
nds.df <- read_nds(site = "Ibahernando", 
                   decor = "Ibahernando",
                   doss = system.file("extdata", 
                                      package = "decorr"))
eds.df <- read_eds(site = "Ibahernando",
                   decor = "Ibahernando",
                   doss = system.file("extdata", 
                                      package = "decorr"))
l_dec_df <- contemp_nds(nds.df, eds.df, selected.nd)
Ibahernando <- plot_dec_grph(nodes = nds.df,
                             edges = eds.df,
                             site = "Ibahernando",
                             decor = "Ibahernando",
                             nd.var = "type",
                             lbl.color = "brown",
                             lbl.size=2.2,
                             doss = system.file("extdata", 
                                                package = "decorr"))
Ibahernando.img <- magick::image_read(Ibahernando)
Ibahernando.contemp <- plot_dec_grph(nodes = l_dec_df[[1]],
                                     edges = l_dec_df[[2]],
                                     site = "Ibahernando",
                                     decor = "Ibahernando",
                                     nd.var = "type",
                                     lbl.color = "brown",
                                     lbl.size=2.2,
                                     doss = system.file("extdata", 
                                                        package = "decorr"))
Ibahernando.contemp.img <- magick::image_read(Ibahernando.contemp)
out <- magick::image_append(c(Ibahernando.img, Ibahernando.contemp.img))
plot(out)
@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> lgrph <- list_dec(imgs, nodes, edges)
R> as.character(named_elements(lgrph[[2]], focus = "edges", nd.var="type"))
## [1] "epee-=-fibule" "bouclier-=-miroir"
## [3] "epee-=-lance" "bouclier-=-chariot_char"
## [5] "chariot_char-+-cheval" "chariot_char-+-cheval#"
## [7] "arc-=-bouclier" "arc-=-chariot_char"
## [9] "arc-=-miroir" "bouclier-=-fibule"
## [11] "bouclier-=-epee" "epee-=-miroir"
## [13] "arc-+-fleche" "chariot_char-+-roue"
## [15] "chariot_char-+-roue#"
R> as.character(named_elements(lgrph[[2]], focus = "nodes", nd.var="type"))
## [1] "chariot_char" "bouclier" "arc" "miroir"
## [5] "fibule" "epee" "lance" "cheval"
## [9] "cheval#" "roue" "roue#" "fleche"
\end{CodeInput}
\end{CodeChunk}

@@

The \pkg{decorr} package has two graphical functions: \code{labels_shadow()} and \code{side_plot()}.

@@

Functions allowing to create a geometric graph for a single decoration are \code{read\_nds()} and \code{read\_eds()}. These functions allow to read respectively a file of nodes and a file of edges (\code{.csv}, \code{.tsv} or \code{.shp} files).

@@

\code{read_nds()} is close to the \proglang{R} native \code{read.table()} function, but allows to read \emph{shapefiles} of nodes. For example, the nodes of the Torrejon Rubio 1 decoration can be read from the the \emph{shapefile}, and the 6\textsuperscript{th} node -- a sword (\code{epee}) -- displayed (Table \ref{tab:nd}).

<<nd, results='asis', echo=TRUE>>=
caption <- "Torrejon Rubio 1 decoration node 6: the sword."
sit <- "Torrejon Rubio" ; dec <- "Torrejon Rubio 1"
nds.df <- read_nds(site = sit, decor = dec, dev = ".shp",
                   doss = system.file("extdata", package = "decorr"))
print(xtable::xtable(nds.df[nds.df$id == 6, ],
                     caption = caption,
                     label = "tab:nd",
                     size = 7),
      include.rownames = FALSE,
      table.placement = "!htbp")
@

@@

\code{read_eds()} permits to read a \code{.csv} or \code{.tsv} dataframe, or a \emph{shapefiles}, of edges. When the data source is a dataframe, the coordinates of the edges are retrieved from the nodes dataframe (Table \ref{tab:ed}).
